BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2014-12-16 Paper/BOAST.tex

The interest in having a robuts and optimally tuned library in this scientific
field is therefore evident.  Techniques are under investigation to provide also
fine-tuned binaries to the end-user rather than the code sources, such that
more agressive interprocedural optimisation can be performed.  This experience
would be therefore a first step toward the release of a tunable optimised
convolution library oriented to computational physics communities.
%   BigDFT implementation of convolution/wavelets is a reference but is limited
%   in scope (1 wavelet family ands its derived operators).  Generalizing to
%   other wavelet families long and difficult if performance are to remain
%   good.
% 
% Idea: \begin{itemize} \item write a generic convolution library in BOAST
% \item BLAS like interface \item Provide a wide variety of building blocks
% \item Optimize thos blocks using BOAST \item Build either from source or
% directly link with binary \item Release complete library \end{itemize}
% 
% 
%   Characteristic use case driven
% 
%   Convolution Library: first step toward a dsl dedicated to wavelet
%   processing for Physics computing.

>>>>

The interest in having a robuts and optimally tuned library in this scientific
field is therefore evident.  Techniques are under investigation to provide also
fine-tuned binaries to the end-user rather than the code sources, such that
more agressive interprocedural optimization can be performed.  This experience
would be therefore a first step toward the release of a tunable optimised
convolution library oriented to computational physics communities.
%   BigDFT implementation of convolution/wavelets is a reference but is limited
%   in scope (1 wavelet family ands its derived operators).  Generalizing to
%   other wavelet families long and difficult if performance are to remain
%   good.
% 
% Idea: \begin{itemize} \item write a generic convolution library in BOAST
% \item BLAS like interface \item Provide a wide variety of building blocks
% \item Optimize thos blocks using BOAST \item Build either from source or
% directly link with binary \item Release complete library \end{itemize}
% 
% 
%   Characteristic use case driven
% 
%   Convolution Library: first step toward a dsl dedicated to wavelet
%   processing for Physics computing.

### 2014-12-16 Paper/specfem.tex
Improved Paper.
Removed useless section.

\section{BOAST Use Cases}

@@

When we started to work on the project (version v2.1 of July 2013) it
supported graphics card GPU acceleration through NVidia \Cuda. This
GPU support comes in addition to the MPI support implemented to enable
parallel computing. Most of \Specfem code-base is written in
\productname{Fortran2003}, only the GPU related parts are written in
C.  The split between CPU and GPU code was done at a rather fine
grain, as the application counted more than 40 GPU kernels. Some of
them were quite simple (\eg performing a few vector operations, but at
massively parallel scale), while at the other side of the spectrum,
some complex kernels took more than 80 parameters and perform very
specific physical transformations.

@@

\subparagraph{Porting kernels to BOAST} Nvidia \Cuda and \OCL are
based on the \emph{same} programming model: a massively parallel
accelerator running in a distinct memory environment. Thanks to that
proximity, we, BOAST developers, have been able to carry out most of
the porting task with only a limited knowledge about \Specfem internal
physics.

@@

Hence, our first milestone in the porting was translating \Specfem
\Cuda kernels to BOAST EDSL. This way, we could ask BOAST framework to
generate CUDA versions of the kernels, plug them back into \Specfem and
get (after fixing compile-time errors---prototypes and naming mistakes
mainly) a first set of \Specfem seismograms.

@@

\subparagraph{Porting runtime to \OCL} The second part of the porting
consisted in translating the CPU-side of the application from \Cuda
API to \OCL API. Most of the functions of the interfaces are very
similar, with only naming-convention and data-structure
distinctions. Hence, it was clear that automatic rewriting tools
(namely \code{sed} regexp and \code{emacs-list} functions) could be
useful. To give an idea of the cost of a \emph{manual} rewriting, we
can count (in \# of \OCL API function calls): 70 kernel ``function
calls'', 790 arguments to set, 230 memory transfers, 160 buffer
creations and 270 releases.

Once all the automatic (and many manual) transformations applied,
compilation errors fixed, and \OCL unsuccessful function calls solved,
the application managed to complete its execution and generated a
first set of seismograms. And again, as expected and feared, these
seismograms were not valid.

As we had already validated BOAST-generated kernels (and trusted \Cuda
and \OCL versions to be semantically identical), we knew that the bugs
were now in the CPU side, and we had to find a way to understand where
\Specfem's \Cuda version of the code diverged from its \OCL
counterpart. To help us in that purpose, we had a strong assumption:
both versions of the code were supposed to perform exactly the same
operations, with the same ``logical'' parameters (the APIs have
\emph{implementation} differences, for instance \OCL has two memory
transfer functions, \code{clEnqueueReadBuffer},
\code{clEnqueueWriteBuffer}, whereas \Cuda has only one, with a
direction parameter \code{cudaMemcpy(..., dir)}, but above that it is
the same functionality nonetheless).

@@

However, in contrast with \code{strace} and \code{ltrace},
\code{GPUTrace} has to be \emph{state-full}. Indeed, most of \OCL and
\Cuda parameters are handles to opaque types, so, in order to provide
more meaningful information, \code{GPUTrace} stores internally
information about these handles (buffer creation size and index,
kernel name and prototype, \etc). % more details here?

Once \code{GPUTrace} was operational, it was easy to confront \Cuda
and \OCL outputs with a graphical \code{diff} tool, and spot the
different porting mistakes: two parameters reversed, an offset
incorrectly applied, \etc. One last problem remained, impossible to
miss because of the seismograms not matching (but with just a slight
drift). We added more verbosity to \code{GPUTrace} output, first the
initial bits of the GPU memory buffers, then their full content. The
drift was visible in the trace, but it was nonetheless unclear where
it started. We finally got it after hours of code review, both on
BOAST kernels and \OCL code: one kernel was \emph{three}-dimensional,
whereas the others were two-dimensional. But for all of them, only two
dimensions were passed. As it was the same person who coded the \OCL
port and the GPU tracer, this third dimension was forgotten in both
side\ldots

\subparagraph{Evaluation} Our \OCL/BOAST port of \Specfem is now
merged in \Specfem's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and x Intel
Xeon processors, we measured similar results between the original
\Cuda version and our BOAST-\Cuda version. With the same set of
optimization, BOAST \Cuda and \OCL version also reported similar
execution time spans. The best execution speed (25\% higher) was
achieved with \Cuda version though, as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed, as of version 1.1, to
\OCL runtime.

By refactoring GPU kernel code in BOAST EDSL, the size of kernel code
shrank by a factor of 1.5 (from xxx LOC to xxx, in part because of
code duplication, but by removing manually unrolled loops). This is
beneficial for \Specfem as it improves the readability and
maintainability of its source code.

We have also been able to enhance \Specfem's non-regression testsuite
by adding per-kernel non-regression tests. This was done with the help
of \code{GPUTrace}, that we used to capture all the input parameters
of a particular \emph{valid} kernel execution, as well as the output
values. Then, during the non-regression testing, BOAST framework can
load these buffer files, allocate and write them on the GPU memory
through \Cuda or \OCL runtime, and trigger the kernel execution. A
comparison of the output values (for instance against a maximal error
level) validates the non-regression.

>>>>

@@

When we started to work on the project (version v2.1 of July 2013) it
supported graphics card GPU acceleration through NVidia \Cuda. This
GPU support came in addition to the MPI support implemented to enable
multi-CPU parallel computing. Most of \Specfem code-base is written in
\productname{Fortran2003}, only the GPU related parts are written in
C.  The split between CPU and GPU code was done at a rather fine
grain, as the application counted more than 40 GPU kernels. Some of
them were quite simple (\eg performing a few vector operations, but at
massively parallel scale), while at the other side of the spectrum,
some complex kernels took more than 80 parameters and perform very
specific physical transformations.

@@

\subparagraph{Porting kernels to BOAST} Nvidia \Cuda and \OCL are
based on the same programming model: a massively parallel accelerator
running in another memory environment. Thanks to that proximity, we,
BOAST developers, have been able to carry out most of the porting task
with only a limited knowledge about \Specfem internal physics.

@@

Hence, our first milestone in the porting was the translation of
\Specfem's \Cuda kernels into BOAST EDSL. This way, we could ask BOAST
framework to generate a CUDA version of the kernels, plug them back
into \Specfem and get (after fixing compile-time errors---prototypes
and naming mistakes mainly) a first set of \Specfem seismograms.

@@

\subparagraph{Porting runtime to \OCL} The second part of the porting
consisted in the translation of the CPU-side of the application, from
\Cuda API to \OCL API. Most of the functions of the interfaces are
very similar, with only naming-convention and data-structure
distinctions. Hence, it was clear that automatic rewriting tools
(namely \code{sed} regexp and \code{emacs-lisp} functions) could be
useful. To give an idea of the cost of a \emph{manual} rewriting, we
can count (in \# of \OCL API function calls): 70 kernel ``function
calls'', 790 arguments to set, 230 memory transfers, 160 buffer
creations and 270 releases.

Once all the automatic (and many manual) transformations were applied,
compilation errors fixed, and \OCL unsuccessful function calls solved,
the application managed to complete its execution and generate a first
set of seismograms. And again, as expected and feared, these
seismograms were not valid.

As we had already validated BOAST-generated kernels (and trusted \Cuda
and \OCL versions to be semantically identical), we knew that the bugs
were now in the CPU side, and we had to find a way to understand where
\Specfem's \Cuda version of the code diverged from its \OCL
counterpart. To help us in that purpose, we had a strong assumption:
both versions of the code were supposed to perform exactly the same
operations, with the same ``logical'' parameters (the APIs have
\emph{implementation} differences, for instance \OCL has two memory
transfer functions, \code{clEnqueueReadBuffer},
\code{clEnqueueWriteBuffer}, whereas \Cuda has only one, with a
direction parameter \code{cudaMemcpy(..., dir)}, but above that, it is
the same functionalities).

@@

However, in contrast with \code{strace} and \code{ltrace},
\code{GPUTrace} has to be \emph{state-full}. Indeed, most of \OCL and
\Cuda parameters are handles to opaque types, so, in order to provide
more meaningful information, \code{GPUTrace} stores internally
information about these handles (buffer creation size and index,
kernel name and prototype, \etc{}). % more details here?

Once \code{GPUTrace} was operational, it was easy to confront \Cuda
and \OCL outputs with a graphical \code{diff} tool, and spot the
different porting mistakes: two parameters reversed, an offset
incorrectly applied, \etc{}. One last problem remained, impossible to
miss because of the seismograms not matching (but with just a slight
drift). We added more verbosity to \code{GPUTrace} output: first the
initial bits of the GPU memory buffers, then their full content. The
drift was visible in the trace, but it was nonetheless unclear where
it started. We finally got it after hours of code review of BOAST
kernels and \OCL code: one kernel was \emph{three}-dimensional,
whereas the others were two-dimensional. But for all of them, only two
dimensions were passed, and one was missing. As it was the same person
who coded the \OCL port and the GPU tracer, this third dimension was
forgotten in both codes\ldots

\subparagraph{Evaluation} Our \OCL/BOAST port of \Specfem is now
merged in \Specfem's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and x Intel
Xeon processors, we measured similar results between the original
\Cuda version and our BOAST-\Cuda version. With the same set of
optimization flags, BOAST \Cuda and \OCL version reported similar
execution time spans. The best execution speed was achieved with \Cuda
version though (25\% higher than \OCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed, to \OCL runtime, as of
version 1.1.

By refactoring GPU kernel code in BOAST EDSL, the size of kernel code
shrank by a factor of 1.5 (from xxx LOC to xxx, partly because of code
duplication, but by removing manually unrolled loops). This is
beneficial for \Specfem as it improves the readability and
maintainability of its source code.

We have also been able to enhance \Specfem's non-regression testsuite
by adding per-kernel non-regression tests. This was done with the help
of \code{GPUTrace}, that we used to capture all the input parameters
of a particular \emph{valid} kernel execution, as well as the output
values. Then, during the non-regression testing, BOAST framework can
load these buffer files, allocate GPU memory and initialize it through
\Cuda or \OCL runtime, and trigger the kernel execution. A comparison
of the output values (for instance against a maximal error level)
validates the non-regression.
